---
title: "Nasal Lavage Fluid Mediator Analysis"
output: html_document
---

# Set up workspace

```{r message = FALSE, warning = FALSE}
# Clear global environment
rm(list=ls())

# Set theme
theme_set(theme_bw())

# Load packages
library(tidyverse) # for data organization and manipulation
library(janitor) # for data cleaning
library(openxlsx) # for reading in files and writing out results
library(ggpubr) # for making Q-Q plots with ggplot 
library(rstatix) # for statistical testing
library(FSA) # package that contains Dunn test function
library(pheatmap) # for heatmaps
library(viridis) # for heatmaps
```

# Import and clean data

Mediator data:
```{r}
nlf_mediator_df_orig <- read.csv("1_InputData/2020_06_11 NLF Data.csv") 

nlf_mediator_df <- nlf_mediator_df_orig %>%
  
  # Omit subject with missing data
  na.omit() %>%
  
  # First name cleaning
  clean_names() %>%
  
  # Convert everything to pg/mL
  mutate_at(vars(contains("ug")), ~ .x*1000000) %>%
  mutate_at(vars(contains("ng")), ~ .x*1000) %>%
  
  # Second column name cleaning step
  rename_with(~gsub(x = .,"_.*","")) %>%
  dplyr::rename("IL-8" = "il", "IgA" = "total", "Neutrophil Elastase" = "neutrophil", "sample_id" = "sample",
         "DEFB4A2" = "defb4a", "DEFB1" = "defb1", "Lactoferrin" = "lactoferrin", "Lysozyme" = "lysozyme")
```

Metadata:
```{r}
metadata <- read.xlsx("1_InputData/ELFmetadata.xlsx") %>%
  rename("sample_id" = "#SampleID") %>%
  filter(sample_id %in% nlf_mediator_df$sample_id)
```

# Test for normality of each mediator

Histograms:
```{r}
nlf_mediator_df %>% pivot_longer(!sample_id, names_to = "variable", values_to = "value") %>%
  ggplot(aes(value)) +
  geom_histogram(fill = "gray40", color = "black", binwidth = function(x) {(max(x) - min(x))/25}) +
  facet_wrap(~ variable, scales = "free", nrow = 2) +
  labs(y = "# of Observations", x = "Value")
```
Q-Q Plots:
```{r}
nlf_mediator_df %>% pivot_longer(!sample_id, names_to = "variable", values_to = "value") %>%
  ggqqplot( x = "value", facet.by = "variable", ggtheme = theme_bw(), scales = "free", nrow = 2)
```
Shapiro-Wilk test:
```{r}
# Apply Shapiro Wilk test
shapiro_res <-  apply(nlf_mediator_df %>% select(-sample_id), 2, shapiro.test)

# Create results data frame
shapiro_res <- do.call(rbind.data.frame, shapiro_res)

# Clean data frame
shapiro_res <- shapiro_res %>% 
  
  ## Add normality conclusion
  mutate(normal = ifelse(p.value < 0.05, F, T)) %>%
  
  ## Remove columns that do not contain informative data
  select(c(p.value, normal)) 

# View cleaned up data frame
shapiro_res
```

# Test for significant differences between groups

```{r}
# Prepare data frame by merging mediator data and metadata
nlf_mediator_df_forstats <- nlf_mediator_df %>%
  left_join(metadata, by = "sample_id")
```


## Between tobacco product use groups

First, perform Kruskal-Wallis test to see if there are overall significant differences in mediator levels between exposure groups.Results demonstrate IL-8 is the only mediator with a close-to-significant overall p-value.
```{r}
# Perform Kruskal-Wallis test
kruskal_res_subjgroup <- nlf_mediator_df_forstats %>%
  dplyr::select(c(DEFB4A2:'IL-8', SubjectGroup)) %>%
  pivot_longer(!SubjectGroup, names_to = "Variable", values_to = "Value") %>%
  group_by(Variable) %>%
  do(tidy(kruskal.test(x = .$Value, g = .$SubjectGroup))) %>%
  dplyr::select(c(Variable, p.value))

# View results
kruskal_res_subjgroup
```
```{r}
# Function to add significant numbers to data frame
signif.num <- function(x) {
  symnum(x, corr = FALSE, na = FALSE, legend = FALSE,
         cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, 1), 
         symbols = c("***", "**", "*", "+", " "))
}

# Function for Dunn test and nicely formatted results
dunn.test <- function(variable, group, data) {
  res <- dunnTest(as.formula(paste0(variable, "~", group)), data = data, method = "bh")
  res <- data.frame(res$res)
  res <- format(res, scientific = FALSE)
  res$AdjSignif <- signif.num(as.numeric(res$P.adj))
  res
}

IL8_dunn_res <- dunn.test("il_8", "subject_group", nlf_mediator_df_forstats %>% clean_names)
```


## Between sexes

```{r warning = FALSE}
# Create results data frame
wilcox_res_sex <- data.frame()

# Define endpoints
mediators <- names(nlf_mediator_df_forstats %>% clean_names %>% select(c(defb4a2:il_8)))

# Iterate through Wilcox test
for (i in 1:length(mediators)) {
  
  # Assign a name to the endpoint variable.
  endpoint <- mediators[i]
  
  # Run wilcox test and store in results data frame.
  res <- wilcox.test(as.formula(paste0(endpoint, "~ sex", sep = "")), 
                     data = nlf_mediator_df_forstats %>% clean_names, 
                     paired = FALSE)
  
  res_df <- data.frame(res$p.value) %>% 
    mutate("variable" = endpoint)
  
  wilcox_res_sex <- rbind(wilcox_res_sex, res_df)
}

# Clean data frame
wilcox_res_sex <- wilcox_res_sex %>% relocate(variable, .before = res.p.value)

# Pull out mediators with significant differences by sex
sex_mediators <- wilcox_res_sex %>%
  filter(res.p.value < 0.05) %>%
  pull(variable)
```

## By sex and exposure

### Normality assessment with log2-transformed data

```{r}
# Log2 transform data
nlf_mediator_df_log2 <- nlf_mediator_df %>%
  mutate(across(where(is.numeric), ~ log2(.x + 1)))

# Plot histograms
nlf_mediator_df_log2 %>% pivot_longer(!sample_id, names_to = "variable", values_to = "value") %>%
  ggplot(aes(value)) +
  geom_histogram(fill = "gray40", color = "black", binwidth = function(x) {(max(x) - min(x))/25}) +
  facet_wrap(~ variable, scales = "free", nrow = 2) +
  labs(y = "# of Observations", x = "Value")
```
```{r}
# Plot Q-Q plots
nlf_mediator_df_log2 %>% pivot_longer(!sample_id, names_to = "variable", values_to = "value") %>%
  ggqqplot( x = "value", facet.by = "variable", ggtheme = theme_bw(), scales = "free", nrow = 2)
```

```{r}
# Apply Shapiro Wilk test
shapiro_res <-  apply(nlf_mediator_df_log2 %>% select(-sample_id), 2, shapiro.test)

# Create results data frame
shapiro_res <- do.call(rbind.data.frame, shapiro_res)

# Clean data frame
shapiro_res <- shapiro_res %>% 
  
  ## Add normality conclusion
  mutate(normal = ifelse(p.value < 0.05, F, T)) %>%
  
  ## Remove columns that do not contain informative data
  select(c(p.value, normal)) 

# View cleaned up data frame
shapiro_res
```

### Two-way ANOVA with log2 data

```{r}
# Add metadata to log2 data frame
nlf_mediator_df_log2 <- nlf_mediator_df_log2 %>%
  left_join(metadata, by = "sample_id") %>%
  clean_names() %>%
  mutate(subject_group = as.factor(subject_group))

# Define endpoints
mediators <- names(nlf_mediator_df_log2 %>% select(c(defb4a2:il_8)))

# Make a data frame where you will store your results
aov_group_sex <- data.frame(Factor = c("subject_group", "sex", "sex:subject_group"))

# Repeat this for loop as many times as there are items in the endpoints vector created above.
for (i in 1:length(mediators)) {
  
  # Assign a name to the endpoint variable.
  endpoint <- mediators[i]
  
  # Run an two-way mixed ANOVA and store results in res.aov
  res.aov <- anova_test(data = nlf_mediator_df_log2, 
                        dv = paste0(endpoint),
                        wid = sample_id,
                        between = c(sex, subject_group)) 
    
  # Extract the results we are interested in.
  res_df <- data.frame(get_anova_table(res.aov)) %>%
    select(c(Effect, p)) %>%
    rename("Factor" = "Effect")
  
  # Rename columns in the results data frame so that the output is more nicely formatted.
  names(res_df)[names(res_df) == 'p'] <- noquote(paste0(endpoint))
  
  # Bind the results to the results data frame.
  aov_group_sex <- merge(aov_group_sex, res_df, by = "Factor", all.y = TRUE)
}
```

Post-hoc testing with Tukey's HSD:
```{r}
# Add combined sex and subject group column
nlf_mediator_df_log2 <- nlf_mediator_df_log2 %>%
  unite(SubjectGroup_Sex, subject_group, sex, sep = "_", remove = FALSE)

# Define endpoints
mediators <- names(nlf_mediator_df_log2 %>% select(c(defb4a2:il_8)))

# Create data frame
pairwise_t_group_sex <- data.frame()

# Repeat this for loop as many times as there are items in the endpoints vector created above.
for (i in 1:length(mediators)) {
  
  # Assign a name to the endpoint variable.
  endpoint <- mediators[i]
  
  # Run an two-way mixed ANOVA and store results in res.aov
  res <- nlf_mediator_df_log2 %>% t_test(as.formula(paste0(endpoint, " ~ SubjectGroup_Sex")))
    
  # Add results to results df
  pairwise_t_group_sex <- rbind(pairwise_t_group_sex, res)
}
```

# Plotting heatmaps for figure panel 1

By group:
```{r}
# Prepare data matrix
nlf_mediator_df_heatmap_group <- nlf_mediator_df_forstats %>%
  mutate(SubjectGroup = as.factor(recode(SubjectGroup, "Smoker" = "SM", "Nonsmoker" = "NS", "Ecig" = "EC"))) %>%
  mutate(SubjectGroup = fct_relevel(SubjectGroup, "NS", "SM", "EC")) %>%
  group_by(SubjectGroup) %>%
  summarise(across(DEFB4A2:'IL-8', ~ mean(.x))) %>%
  column_to_rownames("SubjectGroup") %>%
  select(order(colnames(.))) %>%
  t() 

# Make heatmap
mediator_heatmap_group <- pheatmap(nlf_mediator_df_heatmap_group, 
         color = plasma(100),
         angle_col = c("0"), 
         border_color = "black",
         cellwidth = 30,
         cellheight = 22,
         cluster_cols = FALSE,
         cluster_rows = FALSE,
         fontsize_row = 12,
         scale = 'row',
         fontsize_col = 12)

mediator_heatmap_group

pdf("2_CodeOutput/Heatmap_ByGroup.pdf",
    width = 7, height = 3)
mediator_heatmap_group
invisible(dev.off())
```

By sex:
```{r}
nlf_mediator_df_heatmap_sex <- nlf_mediator_df_forstats %>%
  mutate(Sex = fct_relevel(as.factor(Sex), "Male", "Female")) %>%
  group_by(Sex) %>%
  summarise(across(DEFB4A2:'IL-8', ~ mean(.x))) %>%
  column_to_rownames("Sex") %>%
  select(order(colnames(.))) %>%
  t() 

# Make heatmap
mediator_heatmap_sex <- pheatmap(nlf_mediator_df_heatmap_sex, 
         color = plasma(100),
         angle_col = c("0"), 
         border_color = "black",
         cellwidth = 45,
         cellheight = 22,
         cluster_cols = FALSE,
         cluster_rows = FALSE,
         fontsize_row = 12,
         scale = 'row',
         fontsize_col = 12)

mediator_heatmap_sex

pdf("2_CodeOutput/Heatmap_BySex.pdf",
    width = 7, height = 3)
mediator_heatmap_sex
invisible(dev.off())
```

By sex and group:
```{r}
nlf_mediator_df_heatmap_gs <- nlf_mediator_df_log2 %>%
  dplyr::rename("DEFB4A2" = "defb4a2", "DEFB1" = "defb1", "Lysozyme" = "lysozyme", "Lactoferrin" = "lactoferrin", 
                "IL-8" = "il_8", "Neutrophil Elastase" = "neutrophil_elastase", "IgA" = "ig_a") %>%
  mutate(subject_group = as.factor(recode(subject_group, "Smoker" = "SM", "Nonsmoker" = "NS", "Ecig" = "EC"))) %>%
  mutate(sex = fct_relevel(as.factor(recode(sex, "Male" = "M", "Female" = "F")))) %>%
  unite(SubjectGroup_Sex, subject_group, sex, sep = "_") %>%
  group_by(SubjectGroup_Sex) %>%
  summarise(across(DEFB4A2:'IL-8', ~ mean(.x))) %>%
  column_to_rownames("SubjectGroup_Sex") %>%
  select(order(colnames(.))) %>%
  t() %>% as.data.frame() %>%
  relocate(c(NS_M, NS_F, SM_M, SM_F, EC_M), .before = EC_F)

# Make heatmap
mediator_heatmap_gs <- pheatmap(nlf_mediator_df_heatmap_gs, 
         color = plasma(100),
         angle_col = c("0"), 
         border_color = "black",
         cellwidth = 15,
         cellheight = 22,
         cluster_cols = FALSE,
         cluster_rows = FALSE,
         fontsize_row = 12,
         scale = 'row',
         fontsize_col = 12)

mediator_heatmap_gs

pdf("2_CodeOutput/Heatmap_ByGroupandSex.pdf",
    width = 5, height = 3)
mediator_heatmap_gs
invisible(dev.off())
```



